Discrete and continuous latent states of neural activity in \textit{Caenorhabditis Elegans}

### 2016-12-01 drafts/2016_11_16/writeup.tex
adding hslds code; draft of writeup with results

\section{Introduction}
% The world as it was:
The nematode \celegans~is unique among the model organism of
neuroscience. For decades, we have known the connectome of its 302
neurons, yet a comprehensive understanding of hows its neural activity
reflects sensory processes and produces behavior still eludes
us. While this connectome, or wiring diagram, has provided invaluable
information about the structure of these neural circuits, the critical
missing piece has been large scale functional measurements of these
circuits in action. With the advent of optical recording technologies,
we now possess a mounting set of tools for measuring the activity of large
fractions of these neurons simultaneously. These technologies provide
exciting opportunities to probe the link between neural activity,
sensory processing, and, ultimately, behavior.

@@

% Then one day:
We have developed a probabilistic framework for investigating these questions.
The findings of \citet{kato2015global} --- namely, that neural activity
is low dimensional with behavior-specific dynamics --- suggest that
these multineuronal recordings may be aptly modeled in terms of
a dynamic, low-dimensional latent states. Moreover, they suggest
that neural activity may be parsed into a sequence of discrete
actions, like turning or crawling forward. These two properties are
naturally combined by \emph{switching linear dynamical systems} (SLDS),
a class of probabilistic time series characterized by co-evolving
discrete and continuous latent states.  We have extended this
class of models with a hierarchical Bayesian framework that captures
many of the nuances of these whole-brain recordings. In fitting these
models to recordings from multiple \celegans, we characterize the
globally nonlinear dynamics that govern neural activity, and we
parse these recordings into an interpretable sequence of behavioral segments.

% Raising the stakes:
% Moment of truth:
% World as it is now:
The remainder of this paper is structured as follows. In Section~\ref{sec:slds}
we introduce the class of swithing linear dynamical systems and provide
some insight into the latent states and model parameters. Section~\ref{sec:hslds}
develops a hierarchical extension of the SLDS to share statistical
strength by combining recordings from multiple organisms with partially
overlapping sets of observed neurons.  Then, Section~\ref{sec:results}
presents results of applying these models to a set of recordings from
five separate \celegans. We conclude with directions for further research.

@@

Assume the instantaneous neural activity at time~$t$ for a population
of~$N$ neurons is represented as a vector,~$\by_t \in \reals^N$. In
calcium imaging settings, the entries in this vector may be
instaneous~$\Delta F/F$ measurements, or another signal that captures
neural activity. In this experiment, we use the smoothed time
derivative of~$\Delta F/F$. Over the course of an experiment, we
measure a sequence of vectors, which we combine into a matrix denoted
by~$\by_{1:T}$.

@@

We propose the following hierarchical SLDS (hSLDS) to combine data from
multiple worms. Since these worms are genetically identical, we hypothesize
that their dynamics should be the same. That is, they share the same set
of~$K$ discrete states, as well as the
parameters~$\{\bA^{(k)}, \bb^{(k)}, \bQ^{(k)}, \bpi^{(k)}\}_{k=1}^K$.
We assume the following mapping from latent states to the neural activity
of worm~$w$, time~$t$, and neuron~$n$:
\begin{align}
  y_{w,t,n} &= \gamma_{w,n} \bc_n^\trans \bx_{w,t} + d_{w,n} + \epsilon_{w,t,n}, \\
  \epsilon_{w,t,n} &\sim \distNormal(0, r_{w,n}^2).
\end{align}
Thus, the mapping from latent states to the activity of neuron~$n$ is
partially shared across worms, in that all worms share the
vector~$\bc_n$.  These constitute the rows of an emission
matrix,~$\bC$. Note that, for simplicity, we have required all
discrete states,~$z_t$ to share the same emission matrix. However, we
have introduced a set of worm-specific parameters,~$\{\gamma_{w,n},
d_{w,n}, r_{w,n}\}$. The gain,~$\gamma_{w,n}$, captures differences in
amplitude of the neural signal that may arise from variation in
fluorescent protein expression from worm to worm.  Likewise, the
bias,~$d_{w,n}$, accounts for different baseline levels of
activity. Finally, the noise variance,~$r_{w,n}^2$, enables neuron~$n$
to have different marginal variance from worm to worm, which may arise
due to a number of factors, like protein expression levels and the
clarity of a particular recording.  Figure~\ref{fig:hslds} provides a
graphical representation of this model.

\section{Results}
\label{sec:results}

@@

\section{Bayesian Inference for Switching Linear Dynamical Systems}
\label{app:mcmc}
Our goal is to estimate the posterior probability of a sequence 
of latent states and a set of parameters given the observed data.
From Bayes' rule, we have,
\begin{align}
  p(\bz_{1:T}, \bx_{1:T}, \bTheta \given \by_{1:T}) 
  &= 
  \frac{p(\by_{1:T}, \bx_{1:T}, \bz_{1:T}, \bTheta)}{p(\by_{1:T})}.
\end{align}
The numerator is the joint probability given by Eq.~\eqref{eq:joint}, and
the denominator,~$p(\by_{1:T})$, which is also known as the
\emph{marginal likelihood}, is given by an integral over possible
latent states and parameters,
\begin{align}
  p(\by_{1:T}) &= \int p(\by_{1:T}, \bx_{1:T}, \bz_{1:T}, \bTheta) 
  \, \mathrm{d}\bx_{1:T} \, \mathrm{d}\bz_{1:T} \, \mathrm{d}\bTheta.
\end{align}
Unfortunately, this integral is not efficiently computable for complex
models like the SLDS, forcing us to seek approximate inference methods
instead. Markov chain Monte Carlo (MCMC) methods \todo{cite} offer one
such approach.

To construct our MCMC algorithm, we iteratively sample one set of 
latent states or parameters from its conditional distribution, holding 
the rest fixed, in a technique known as Gibbs sampling \todo{cite}. 
There are five main sets of parameters to sample, detailed below.
\begin{enumerate}
  \item \textit{Gibbs sampling the discrete latent states,~$\bz_{1:T}$:}

>>>>

\section{Introduction}
% The world as it was:
The nematode \celegans~is unique among the model organism of
neuroscience. For decades, we have known the connectome of its 302
neurons, yet a comprehensive understanding of hows its neural activity
reflects sensory processes and produces behavior still eludes
us \citep{white1986structure}. While this connectome, or wiring diagram, has provided invaluable
information about the structure of these neural circuits, the critical
missing piece has been large scale functional measurements of these
circuits in action. With the advent of optical recording technologies,
we now possess a mounting set of tools for measuring the activity of large
fractions of these neurons simultaneously. These technologies provide
exciting opportunities to probe the link between neural activity,
sensory processing, and, ultimately, behavior.

@@

% Then one day:
We have developed a probabilistic framework for investigating these questions.
The findings of \citet{kato2015global} --- namely, that neural activity
is low dimensional with behavior-specific dynamics --- suggest that
these multi-neuronal recordings may be aptly modeled in terms of
a dynamic, low-dimensional latent states. Moreover, they suggest
that neural activity may be parsed into a sequence of discrete
actions, like turning or crawling forward. These two properties are
naturally combined by \emph{switching linear dynamical systems} (SLDS),
a class of probabilistic time series characterized by co-evolving
discrete and continuous latent states.  We have extended this
class of models with a hierarchical Bayesian framework that captures
many of the nuances of these whole-brain recordings. In fitting these
models to recordings from multiple \celegans, we characterize the
globally nonlinear dynamics that govern neural activity, and we
parse these recordings into an interpretable sequence of behavioral segments.

% Raising the stakes:
% Moment of truth:
% World as it is now:
The remainder of this paper is structured as follows. In Section~\ref{sec:slds}
we introduce the class of switching linear dynamical systems and provide
some insight into the latent states and model parameters. Section~\ref{sec:hslds}
develops a hierarchical extension of the SLDS to share statistical
strength by combining recordings from multiple organisms with partially
overlapping sets of observed neurons.  Then, Section~\ref{sec:results}
presents results of applying these models to a set of recordings from
five separate \celegans. We conclude with directions for further research.

@@

The switching linear dynamical system (SLDS) is a powerful model for
stochastic time series with nonlinear dynamics \citep{ackerson1970state, chang1978state,
  hamilton1990analysis, bar1993estimation, ghahramani1996switching,
  murphy1998switching, fox2009nonparametric}, and as such, it is
particularly well suited to the analysis of whole-brain activity.
Assume the instantaneous neural activity at time~$t$ for a population
of~$N$ neurons is represented as a vector,~$\by_t \in \reals^N$. In
calcium imaging settings, the entries in this vector may be
instantaneous~$\Delta F/F$ measurements, or another signal that captures
neural activity. In this experiment, we use the smoothed time
derivative of~$\Delta F/F$. Over the course of an experiment, we
measure a sequence of vectors, which we combine into a matrix denoted
by~$\by_{1:T}$.

@@

We propose the following hierarchical SLDS (hSLDS) to combine data from
multiple worms. Since these worms are genetically identical, we hypothesize
that their dynamics should be the same. That is, they share the same set
of~$K$ discrete states, as well as the
parameters~$\{\bA^{(k)}, \bb^{(k)}, \bQ^{(k)}, \bpi^{(k)}\}_{k=1}^K$.
We assume the following mapping from latent states to the neural activity
of worm~$w$, time~$t$, and neuron~$n$:
\begin{align}
  y_{w,t,n} &= \gamma_{w,n} (\bc_n^\trans \bx_{w,t} + d_{n}) + \epsilon_{w,t,n}, \\
  \epsilon_{w,t,n} &\sim \distNormal(0, r_{w,n}^2).
\end{align}
Thus, the mapping from latent states to the activity of neuron~$n$ is
partially shared across worms, in that all worms share the
vector~$\bc_n$.  These constitute the rows of an emission
matrix,~$\bC$. Note that, for simplicity, we have required all
discrete states,~$z_t$ to share the same emission matrix. However, we
have introduced a set of worm-specific, scalar
parameters,~$\{\gamma_{w,n}$ and~$r_{w,n}\}$. The
gain,~$\gamma_{w,n}$, captures differences in amplitude of the neural
signal that may arise from variation in fluorescent protein expression
from worm to worm.  The noise variance,~$r_{w,n}^2$, enables
neuron~$n$ to have different marginal variance from worm to worm,
which may arise due to a number of factors, like protein expression
levels and the clarity of a particular recording.  Whereas the general
SLDS allows for covariance in the noise from neuron to neuron, for
simplicity we have assumed that this matrix is worm-specific and
diagonal. Figure~\ref{fig:hslds} provides a graphical representation
of this model.

% Talk about the prior on gain... we used a truncated normal prior,
% maybe the fact that we needed such a constrained prior means that
% this gain parameter is a bad idea after all...

\section{Results}
\label{sec:results}
\begin{figure}[t]
\centering%
\includegraphics[width=5.5in]{shared_activity_worm1.pdf} 
\caption{Fluorescence traces of 15 neurons in the second worm. These
  neurons were identified in all 5 worms. Axis limits are the same for
  all neurons. }
\label{fig:shared_activity}
\end{figure}

We fit the hSLDS to recordings of five \textit{C. Elegans} collected
by \cite{kato2015global}. Each recording is 18 minutes long with a
sampling frequency of 3Hz, yield 3240 time steps per worm.  The
recordings contain~$N_w=109$, $107$, $131$, $126$, and $129$ neurons,
respectively. Of these, a subset have been manually identified.
Figure~\ref{fig:identified_neurons} shows the neurons that were
identified in each worm. We will capitalize on those neurons that are
identified in multiple worms in order to learn a shared subspace of
neural activity and a shared set of dynamical states for all
\textit{C. Elegans}. Figure~\ref{fig:shared_activity} shows the neural
activity of 15 identified neurons in the second worm. We immediately
recognize repeating patterns of activity; these will correspond to
characteristic ``loops'' in the low dimensional portrait of the
population activity.

\paragraph{Neural Activity is Low Dimensional}

% PCA trajectories of neural activity
\begin{figure}[t]
  \centering%
  \begin{subfigure}[b]{.19\linewidth}
    \includegraphics[width=\textwidth]{pca_trajectory_worm0.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{.19\linewidth}
    \includegraphics[width=\textwidth]{pca_trajectory_worm1.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{.19\linewidth}
    \includegraphics[width=\textwidth]{pca_trajectory_worm2.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{.19\linewidth}
    \includegraphics[width=\textwidth]{pca_trajectory_worm3.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{.19\linewidth}
    \includegraphics[width=\textwidth]{pca_trajectory_worm4.pdf}
  \end{subfigure} 
  \caption{Principal components analysis of each individual worm.
    In all cases, the neural activity is relatively low dimensional,
    with large percentages of the variance captured by the first three
    principal components (PC's). Above, the projection of roughly 100-dimensional
    neural activity onto the subspace spanned by the first three PC's.
    The hierarchical SLDS will find a common subspace for all worms, and,
    furthermore, segment the low dimensional trajectories into a sequence
    of discrete pieces.}
\label{fig:pca_trajectory}
\end{figure}

A fundamental assumption of the hSLDS is that neural activity is
low dimensional. To test this assumption, we first run principal
components analysis (PCA) on each worm's
population activity (a~$T \times N_w$ matrix) separately. We find
that the top three principal components (PC's)capture a large fraction
of the variance for all worms, as shown in the bottom panels of
Figure~\ref{fig:pca_trajectory}. The top panels show the projection
of neural activity onto the first three PC's. While the PCA
projections show qualitatively similar patterns, like intersecting
and orthogonal loops, the subspaces are not aligned across worms
and the trajectories are not particularly smooth or interpretable.
The hSLDS addresses these issues by learning a shared subspace
across all worms, and segmenting the low dimensional
trajectories into a sequence of states with simple, linear dynamics.

\paragraph{The hSLDS reveals a shared subspace and canonical dynamics across worms}

% Inferred dynamical system -- set of discrete latent states
\begin{figure}[t]
  \centering%
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{xs_3d_worm0.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{xs_3d_worm1.pdf}
  \end{subfigure}
  \\
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{xs_3d_worm2.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{xs_3d_worm3.pdf}
  \end{subfigure}  
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{xs_3d_worm4.pdf}
  \end{subfigure}
  \caption{Latent states discovered by the hierarchical SLDS.
    Since the emission matrix,~$\bC$, is shared across worms, the
    coordinates of the latent states are shared as well. This
    reveals a common pattern of low dimensional activity characterized
    by interlocking and opposing loops. Colors denote the discrete
    latent state inferred for each part of the trajectory. For example,
    one loop begins with the latent state shooting out from the origin
    along the orange and purple states, and then returning via the blue
    state. Another begins with the green state, which swings upward and
    is then followed by a return via the red state. In worms 3 and 5,
    the upward swing is slightly tilted, as captured by the pink 
    rather than the green state. While harder to see in this view, there
    is a third loop in yellow that lies in the~$(x_1, x_2)$ plane.
    See supplementary videos for further detail.
  }
\label{fig:hslds_trajectory}
\end{figure}

The hSLDS shares information across worms in order to learn a
canonical subspace of neural activity and a common set of
low-dimensional dynamics. Figure~\ref{fig:hslds_trajectory} shows the
inferred continuous latent states,~$\bx_{1:T}$, of the hSLDS for each
of the five worms. There are three points to note, First, in contrast
to the PCA trajectories in Figure~\ref{fig:pca_trajectory}, the
similarity in hSLDS trajectories is readily apparent. This is due to
the fact that all worms share the same subspace, which is defined by the
global emission matrix~$\bC$. Second, the hSLDS trajectories
are considerably smoother than their PCA counterparts. This is a
direct result of the set of linear dynamics that comprise the hSLDS.
Assuming that the latent state,~$\bx_{t+1}$, is linearly related to the
preceding state,~$\bx_t$, induces an inductive bias toward smooth
trajectories in latent space. This reduces the discontinuities and noise
in the PCA trajectories and reveals the characteristic patterns of
latent dynamics. Third, the hSLDS trajectories have been segmented into
snippets, each corresponding to a different discrete latent state,~$z_t$, with
its own set of linear dynamics. These discrete states illustrated
as different colors in the hSLDS trajectories. These provide a
handle into understanding the latent dynamics of \textit{C. Elegans}
population activity, and a useful latent variable for relating
activity to behavior.

\paragraph{Neural dynamics switch between a set of linear regimes}
% 3D dynamics
\begin{figure}[th!]
  \centering%
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{dynamics_3d_0.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{dynamics_3d_1.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{dynamics_3d_2.pdf}
  \end{subfigure}
  \\
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{dynamics_3d_3.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{dynamics_3d_4.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{dynamics_3d_5.pdf}
  \end{subfigure}
  \\
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{dynamics_3d_6.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{dynamics_3d_7.pdf}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\linewidth}
    \includegraphics[width=\textwidth]{dynamics_3d_8.pdf}
  \end{subfigure}  
%  \\
%  \begin{subfigure}[b]{0.32\linewidth}
%    \includegraphics[width=\textwidth]{dynamics_3d_9.pdf}
%  \end{subfigure}
%  \begin{subfigure}[b]{0.32\linewidth}
%    \includegraphics[width=\textwidth]{dynamics_3d_10.pdf}
%  \end{subfigure}
%  \begin{subfigure}[b]{0.32\linewidth}
%    \includegraphics[width=\textwidth]{dynamics_3d_11.pdf}
%  \end{subfigure}  
  \caption{The discrete dynamical states that compose this switching
    linear dynamical system. The vector fields point in the direction of
    ~$\bA^{(k)} \bx + \bb^{(k)}$~($k=1 \ldots 9$). While the dynamics are
    defined for all~$\bx$,
    they are only plotted at locations where the corresponding state is used.
    States are ordered from most common (1)
    to least common (9); see supplementary material for remainder of
    states. The most common states (1 and 2) simply decay to the
    origin. State 3 is a tight loop in the~$(x_1, x_2)$ plane. States
    4 and 8 are larger loops in the~$(x_2, x_3)$ plane. States 5, 6,
    and 9 are sections of a third loop opposite to State 3. Finally,
    State 7 seems to bridge between loops.}
\label{fig:dynamics}
\end{figure}

As we have stated, the hSLDS segments the low dimensional trajectories
into a sequence of discrete states, each governed by simple linear dynamics.
These dynamics are shown in Figure~\ref{fig:dynamics}. The states are ordered
from most common (State 1) to least common (State 9). The model was fit with~$K=15$
states, but only 11 were used. The remaining two are omitted from this figure
since they are used very rarely. The vector fields shown here point in the
direction of~$\bA^{(k)} \bx_t + \bb^{(k)}$ for~$k=1, \ldots, 9$. While these
dynamics are defined for all~$\bx_t \in \reals^3$, for clarity, they are only plotted
at the locations in latent space where the corresponding state is deployed.

The most common states (1 and 2) correspond to decays toward the origin.
This is to be expected, since the neurons are quiescent most of the time
(c.f. Figure~\ref{fig:shared_activity}). The remaining states capture
different sections of the intersecting loops seen in Figure~\ref{fig:hslds_trajectory}.
State 3 is a short loop to the left in the~$(x_1, x_2)$ plane. States 4
and 8 correspond to the large upward swings that trace out loops in the~$(x_2, x_3)$
plane. States 5, 6, and 9 correspond to loops that swing out to the right
of the origin, again in the~$(x_1, x_2)$ plane. Finally, state 7 seems to
connect the~$(x_2, x_3)$ loop to the~$(x_1, x_2)$ loop, bypassing the origin.

This discrete segmentation of the latent state trajectories serves multiple
purposes. First, by stitching together discrete states with unique
linear dynamics, the hSLDS models a globally \emph{nonlinear} dynamical
system.  Second, these discrete latent states provide a valuable covariate
for understanding switches in the neural activity. As we will see below,
when we condition on the entry into these discrete states, characteristic
patterns of population activity are revealed. Third, these states provide
an interpretable summary of the complex, global brain activity. Next, we
will show that these discrete states correspond to meaningful behaviors.

\paragraph{Discrete states correspond to distinct behaviors}

% Discrete state transitions mimic those of hand-labeled behavioral segments
\begin{figure}[t]
\centering%
\includegraphics[width=5.5in]{discrete_state_cps_worm1.pdf} 
\caption{The inferred discrete states align with the manual segmentation
  of \citet{kato2015global}. \textit{Top:} samples from the posterior
  distribution of discrete states,~$z_{1:T}$. Each row is a sample, and
  each color denotes a distinct value from 1 to~$K=15$. \textit{Middle:}
  from these samples we estimate the probability of a change-point
  (i.e.~$z_t \neq z_{t-1}$). \textit{Bottom:} The inferred change-points
  align with the manual segmentation of \citet{kato2015global}. Note,
  however, that the colors do not perfectly align since there is not a
  one-to-one correspondence between the inferred and manual states. For
  example, the blue and red states of~\citet{kato2015global}, which
  correspond to ``REVSUS'' and ``SLOW,'' respectively, are combined into
  inferred State 1.
}
\label{fig:discrete_state_comparison}
\end{figure}

\citet{kato2015global} manually labeled segments of time corresponding to
distinct behaviors, like forward or reverse crawling and dorsal or ventral
turns. This process is both laborious and subjective. The hSLDS provides
a principled and automated alternative to manual labeling, by automatically
segmenting activity into snippets governed by linear dynamics. We find that
the manual and hSLDS segmentations are closely aligned.

Figure~\ref{fig:discrete_state_comparison} illustrates both the uncertainty
of the discrete labeling as well as the alignment with the ``ground truth''
labels. The top panel shows samples from the posterior distribution over
latent state sequences. Each row corresponds to a sample of~$z_{1:T}$, and
as before, the colors denote different discrete states. The variability
from sample to sample highlights the uncertainty that is inherent in assigning
a discrete label to a point in time --- where one state ends and the next
begins is not entirely clear from the noisy data.

From this sequence of discrete state samples we derive an estimate of the
\emph{change-point probability}, i.e. the probability that~$z_t \neq z_{t-1}$.
This is shown in the middle panel. In some instants it is eminently clear
that the discrete state has switched; this is due to large changes in the
dynamical trajectory that favor one state over another. More often, however,
this probability is lower and distributed over a window of time. Thus, our
Bayesian approach provides a principled means of assessing this uncertainty.

% Neural reconstruction
\begin{figure}[t]
\centering%
\includegraphics[width=5.5in]{neural_activity_worm_1.pdf} 
\caption{Observed (black) and smoothed (blue) activity of six neurons
  from worm 2. The smoothed activity is given by~$\gamma_{w,n}
  (\bc_n^\trans \bbE[\bx_{w,t}] + d_n)$. Shading denotes ~$\pm 1$
  standard deviation under the posterior given by~$\sqrt{\gamma_{w,n}^2 \bc_n^\trans
    \text{Var}[\bx_{w,t}] \bc_n + \sigma^2_{w,n}}$. Inferred discrete states
  are shown above for reference. }
\label{fig:neural_reconstruction}
\end{figure}

Finally, we see that these change-points align with the manual segmentation
of~\citet{kato2015global}, which is shown in the bottom panel. Note that the
manual and inferred state labels are not in one-to-one correspondence. This
is to be expected given both the invariance to permutation (though both are
sorted by usage), but more importantly because some manual states are grouped
into one inferred state. For example, the red and blue manual states, which
are labeled ``REVSUS'' and ``SLOW'', respectively, are combined in the blue
inferred state. Likewise, the manual sequence of purple followed by green,
which corresponds to a dorsal turn followed by a forward crawl, is combined
in the yellow inferred state.
Other states, like the yellow manual state, which corresponds
to a ventral turn, are separated into two inferred states, here red
and green. While the exact divisions between states are somewhat arbitrary,
the change points in these trajectories are clearly revealed by the hSLDS.

\paragraph{The hSLDS captures salient patterns of neural activity}

% Reconstruct the neural activity
As a generative model, the hSLDS provides a distribution over observed
neural activity. This provides a valuable sanity check for our results ---
does the model capture the variability of the data?
Figure~\ref{fig:neural_reconstruction} shows that indeed it does. We
plot the observed traces of six neurons in black and the predicted traces
under the posterior of the hSLDS in blue. With a Bayesian approach, we
also obtain an estimate of the uncertainty of these predictions, shown
here as shaded blue regions around the mean. We see that for most neurons,
the predicted and observed trajectories are well aligned, indicating that
the hSLDS with~$D=3$ dimensional latent states has sufficient capacity
to capture the variability of the data; however, some spikes in the activity
of neuron AVBL are smoothed over in this reconstruction.

% Where do neurons lie in latent space?
% - which neurons are most interesting?
% - any listed in their paper
\paragraph{The hSLDS discovers a low-dimensional embedding of neurons}
\begin{figure}[t]
\centering%
\includegraphics[width=4.in]{neuron_embedding.pdf} 
\caption{Three-dimensional embedding of neurons given by the
  rows of the emission matrix,~$\bc_n$. For reference, the
  true, one-dimensional location of the soma is shown above
  (jittered for clarity). While most neurons are clustered
  near the origin (reflecting the fact that most neurons
  have low amplitude activity), the more active neurons (shown
  above) stand out.}
\label{fig:neuron_embedding}
\end{figure}

In addition to yielding an interpretable, low-dimensional portrait
of neural activity, the hSLDS also provides an embedding of neurons
in latent space via the rows of the emission matrix,~$\bC$. Nearby
neurons in latent space will have highly correlated activity, and neurons
that are opposed will be anti-correlated. Figure~\ref{fig:neuron_embedding}
shows this embedding for the 60 neurons that are identified in at least
one of the five worms. For reference, the true one-dimensional location
of the neuron's soma along the head to tail axis is shown above.
Most of the neurons are clustered around the origin, reflecting the fact
that most neurons have little activity. However, two pairs of neurons
stand out: (AVAL, AVAR) and (RIML, RIMR). These pairs of neurons have larger
amplitude activity as shown above. These pairs both drive backward locomotion
(reversal), and are correlated via a gap junction.

% State entry triggered neural responses
\paragraph{Discrete state onset reveal stereotyped population activity}

\begin{figure}[t]
  \centering%
  \begin{subfigure}[b]{5.5in}
    \includegraphics[width=\textwidth]{neural_responses_subset_0.pdf}
  \end{subfigure} \\
  \begin{subfigure}[b]{5.5in}
    \includegraphics[width=\textwidth]{neural_responses_subset_1.pdf}
  \end{subfigure} \\
  \begin{subfigure}[b]{5.5in}
    \includegraphics[width=\textwidth]{neural_responses_subset_2.pdf}
  \end{subfigure}\\
  \begin{subfigure}[b]{5.5in}
    \includegraphics[width=\textwidth]{neural_responses_subset_3.pdf}
  \end{subfigure}
  \caption{Neural activity conditioned upon entry into each discrete
    state. The~$x$-axis is time since entering the discrete state
    (from 0 to 5 seconds), and the $y$-axis is the average neural
    response~$\pm 2$ standard deviations.  Light gray traces show a
    sample of individual neural activity traces.  Each row corresponds
    to the discrete state with the corresponding color. Here shown are
    States 1-4.}
  \label{fig:state_trigged_activity_1}
\end{figure}

\begin{figure}[t]
  \centering%
  \begin{subfigure}[b]{5.5in}
    \includegraphics[width=\textwidth]{neural_responses_subset_4.pdf}
  \end{subfigure} \\
  \begin{subfigure}[b]{5.5in}
    \includegraphics[width=\textwidth]{neural_responses_subset_5.pdf}
  \end{subfigure} \\
  \begin{subfigure}[b]{5.5in}
    \includegraphics[width=\textwidth]{neural_responses_subset_6.pdf}
  \end{subfigure}\\
  \begin{subfigure}[b]{5.5in}
    \includegraphics[width=\textwidth]{neural_responses_subset_7.pdf}
  \end{subfigure}
  \caption{Same as figure above but for States 5-8.}
  \label{fig:state_trigged_activity_2}
\end{figure}

The discrete latent states partition neural activity into interpretable
and behaviorally relevant segments. As such, they are natural covariates
for neural activity. Figures~\ref{fig:state_trigged_activity_1}
and~\ref{fig:state_trigged_activity_2} illustrate the distribution of
neural responses upon entry into each of the top 8 discrete states.
The~$x$ axis in these plots denotes time from 0 to 5
seconds after entry into the corresponding state. The colored line
denotes the mean response of the corresponding neuron (same $y$ limits
in all panels), and the shaded regions denote~$\pm 2$ standard deviations.
The light gray traces show a sample of randomly chosen neural responses.
That AVA and RIM neurons are associated with reverse crawling suggests
that states 2, 3, and 8 are related to forward locomotion (since the neural
activity is decreasing) and states 5, 6, and 7 are related to the
reverse locomotion (since the activity is increasing).

% Residual analysis -- how well does the model fit?
% - baselines: probabilistic PCA, classical SLDS
% - note that these are less constrained!

\section{Discussion}
The elucidation of global brain dynamics and their relation to behavior
is perhaps the foremost goal of systems neuroscience. \celegans, with
its well-studied connectome and corresponding array of optical imaging tools,
provides many exciting opportunities to make progress toward this goal.
To do so, however, requires statistical methods capable of revealing latent
structure in complex neural data, and combining data across many worms.
The hierarchical SLDS satisfies many of these demands, decomposing
high-dimensional, nonlinear dynamics into a sequence of discrete latent
states with low-dimensional, linear dynamics. Moreover, by sharing across
worms while allowing for individual variability, the hSLDS reveals a
common set of latent dynamics.

% Remaining questions
Many questions remain. First, as a generative model, the hSLDS falls
somewhat short with its relatively limited model for discrete dynamics.
It is clear from these portraits of neural activity that discrete states
are deployed in very localized regions of latent space. Our recent work on
recurrent switching linear dynamical systems (rSLDS) leverages this type
of structure \citep{linderman2016recurrent}. In future work, we intend to
combine the hSLDS and rSLDS and apply them to this data.

Second, our analyses of state-triggered neural activity have only scratched
the surface. We intend to look more closely at the population responses
in these distinct latent states and search for further insights into how
this low-dimensional population activity drives behavior.

Third, we have not leveraged the availability of the \celegans connectome
in this work. In theory, this connectome should provide strong information
about the correlational structure of population activity. It remains an
open question how best to combine this anatomical data with state space
models like these.

Finally, while we have utilized the neurons that are found in multiple
worms to learn a shared subspace of activity and a common set of discrete
states, we could further improve upon this work by inferring neuron
identities for the hundreds of unlabeled neurons.  Their rough locations
and the correlations across worms provides a substantial amount of information
to tackle this problem.

@@

\section{Bayesian Inference for Switching Linear Dynamical Systems}
\label{app:mcmc}
Our goal is to estimate the posterior probability of a sequence 
of latent states and a set of parameters given the observed data.
From Bayes' rule, we have,
\begin{align}
  p(\bz_{1:T}, \bx_{1:T}, \bTheta \given \by_{1:T}) 
  &= 
  \frac{p(\by_{1:T}, \bx_{1:T}, \bz_{1:T}, \bTheta)}{p(\by_{1:T})}.
\end{align}
The numerator is the joint probability given by Eq.~\eqref{eq:joint}, and
the denominator,~$p(\by_{1:T})$, which is also known as the
\emph{marginal likelihood}, is given by an integral over possible
latent states and parameters,
\begin{align}
  p(\by_{1:T}) &= \int p(\by_{1:T}, \bx_{1:T}, \bz_{1:T}, \bTheta) 
  \, \mathrm{d}\bx_{1:T} \, \mathrm{d}\bz_{1:T} \, \mathrm{d}\bTheta.
\end{align}
Unfortunately, this integral is not efficiently computable for complex
models like the SLDS, forcing us to seek approximate inference methods
instead. Markov chain Monte Carlo (MCMC) methods \citep{gilks2005markov, robert2013monte} offer one
such approach.

To construct our MCMC algorithm, we iteratively sample one set of 
latent states or parameters from its conditional distribution, holding 
the rest fixed, in a technique known as Gibbs sampling. 
There are five main sets of parameters to sample, detailed below.
\begin{enumerate}
  \item \textit{Gibbs sampling the discrete latent states,~$\bz_{1:T}$:}
